On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-02 manuscripts/PNAS/review/point_by_point_response.txt

-- The paper does a good job of motivating the general problem of sparse coding to a general audience. In particular, there is a good review of the areas in which dictionary learning arises.

-- I think the biggest potential issue for publication in PNAS is that the significance of the result needs more discussion. Specifically, the authors could address why going from the noise-free case to the noisy case is not incremental. PNAS specifically targets submissions that go beyond what could be published in a more specialized journal. Could the authors, for example, contrast the proof technique between the noisy and noise-free cases? Or, perhaps, the practical implications? 
-----------------

RESPONSE:

With all due respect to the reviewer, we believe the practical implications of a noisy result over a noiseless result are plainly evident: never has there been an experiment without noise and an unstable model is therefore useless in practice. It seems to us that the reviewer’s fundamental concern is more likely that our proof may amount to too straight-forward of an extension of the approach taken in [Hillar & Sommer (2015)] to satisfy the criterion for publication in PNAS. We believe this concern stems from two misconceptions the reviewer seems to us to have: that our main result (Thm. 1) is merely an extension of that in [Hillar & Sommer (2015)] to the noisy case, and that our proof technique is only an incremental tweak on the approach taken in [Hillar & Sommer (2015)].

It is understandable from the presentation of our proof of our main result (Thm. 1) that our approach appears on the surface to be a straight-forward extension of that taken in [Hillar & Sommer (2015)]. As observed by our second reviewer, however, the extension to the noisy case did indeed require an original combination of several results in the literature. Aside from these required modifications, our proof of (Thm. 1) differs from the approach taken by [Hillar & Sommer (2015)] in a key way that is tucked into the proof of our Main Lemma — more specifically, by the addition of Lemma 5. [More detail on what goes down in Lemma 5?] We have updated the manuscript to make the reader more keenly aware of this. To be clear, this new combinatorial perspective we take on the problem yields the following powerful new conclusions beyond those in [Hillar & Sommer (2015)].

@@

2) An extension to the case where the number of dictionary elements is unknown: The results of [Hillar & Sommer (2015)] only apply to the case where the matrix B has the same number of columns as A. We forgo this assumption and show that B must have at least as many columns as A and contains (up to noise) a subset of the columns of A. The size of this subset depends on a simple relation between the number of columns of B and the regularity of the support-set hypergraph.

3) No spark condition requirement for the generating matrix A: One of the mathematically significant contributions made by [Hillar & Sommer (2015)] was to forgo the constraint that the recovery matrix B also satisfy the spark condition, in contrast to all previously known uniqueness results which (either explicitly or implicitly) made this assumption. Our proof is powerful enough to show that, in fact, even the matrix A need not satisfy the spark condition to be identifiable! Rather, it need only be injective on the union of subspaces with supports that form sets in a hypergraph satisfying the SIP. We had omitted this point in our original submission of the manuscript to keep things simple, but have decided to include this interesting fact in our resubmission given that a recently approved NSF grant _____ makes use of this fact in its description of methods. This also required us to redefine the restricted matrix lower bound L to be in terms of a hypergraph, which is an interesting object for further study in its own right.

[4) Implications for the computational complexity of the dictionary learning problem. If the size of the support set of reconstructing codes is polynomial in m,k then the pigeonholing argument in the proof of Thm. 1 requires only a polynomial number of samples distributed over a polynomial number of supports; thus N is polynomial in m, k. [Chris? How does this have implications..?] This point was only hinted at in the Discussion section of our original submission, but we have included it in the updated manuscript to make clear the power of our approach over that taken in [Hillar & Sommer (2015)].

@@

Let A satisfy eqn [2], x_1,...,x_N be a set of k-sparse vectors, and E \subseteq \choose{[m],k}, and \bar{m} be given and satisfy the following property:

For every S \in E, there exists a subset of x_1,...,x_N of at least size (k-1)\choose{\bar{m},k} with support in S. Moreover, any k vectors in this subset are linearly independent.

Is this what you meant? 
-----------------

RESPONSE: Yes, though actually for every S \in E there are *more than* (k-1)\choose{\bar{m},k} vectors with support in S (not “at least”). Thank you for helping us to clarify our statements. We have updated the manuscript to disambiguate these conditions.

@@

RESPONSE: Yes, there is indeed. While all existing uniqueness results require a sufficient number of samples be drawn from *each* of {m \choose k} possible supports, we have shown that in fact it is enough that a sufficient number of samples be drawn from every support in a subset of drastically smaller size, provided this subset forms a regular hypergraph satisfying the singleton intersection property. As was pointed out in our manuscript, a support set of size m satisfying these criteria can always be constructed for any k < m (e.g. take the consecutive intervals of length k in some cyclic order on [m]), and in certain cases such a set can be constructed from as few as 2\sqrt(m) supports (e.g. when m = k^2, take the supports to be the rows and columns of [m] arranged in a square grid).

-----------------
In these terms, it is also unclear to me if a theoretical extension of the uniqueness results to the stability results merit publication in PNAS given the "top 10%" and "sufficient general interest" requirements for publication. It definitely is deserving of publication in a specialty journal.

The work presented is analytical and does not rely on data validation. 
-----------------

RESPONSE: We have addressed this concern in our response to the first reviewer’s identical comment. It seems both reviewers haven’t noticed that our results generalize the results of [Hillar & Sommer (2015)] far beyond a straight-forward extension to the noisy case.

@@

-----------------
In the paragraph "Since f is an analytic function..." (page 3) it is argued that a specific construction of a suitable matrix implies that "almost every real nxm matrix... satisfies Eq. (2)". This appears to overstate the result, in my opinion - can the authors elaborate on why "almost every real matrix" can be obtained from this construction based on a Vandermonde matrix with additional zero rows? 
-----------------

RESPONSE: The point we make here is not that almost every real matrix satisfying Eq. (2) can be obtained from the construction based on a Vandermonde matrix which we provide; rather, the point is that almost every matrix satisfies Eq. (2) because at least one matrix does (e.g. our construction) and therefore, since f is analytic, the set of all those that don’t satisfy the condition has measure zero. We have updated the manuscript to make this argument more clear.

@@

-----------------
Some typos appear in Eq. (24), closing } missing from the denominators of the second and third inequalities. Additionally, it was not clear to me how B_{pi(S)} change to B_{pi(T)}. 
-----------------

RESPONSE: Thank you for pointing out this typo in the second and third inequalities of Eq. (24). We have also clarified in our resubmission how it is that B_{pi(S)} can be replaced with B_{pi(T)} (it is because d(U, V) \leq d(U’,V) whenever U \subseteq U’).

### 2017-08-03 manuscripts/PNAS/review/point_by_point_response.txt

RESPONSE: We have included in our revised manuscript a more compelling exposition of the significance of our results and their relevance to the general scientific audience, and thank the editor for the opportunity to address this concern shared by both reviewers. [Summarize high level major points in our cover here.  Also, mention we try to flesh this out in the next comment below.]

-----------------	
Since the result is an extension of the noise-free result of [Hillar & Sommer (2015)] to a noisy case, the authors should provide better justification for the significance of contributions. The main result seems to be based on an approach similar to [Hillar & Sommer (2015)], and there is little to no discussion of the computability and tightness of the sufficient condition for stability in the main results. Thus, the submission does not seem to be greatly more valuable than [Hillar & Sommer (2015)] and also not conclusive regarding the central problems of sparse linear coding and dictionary learning. That being said, the paper (with the modifications below) would certainly merit publication in very solid journal like IEEE Transactions on Information Theory, where it would be a median paper, in my opinion.  
-----------------

RESPONSE:

We agree with the reviewer that a straight-forward extension of the noise-free result of [Hillar & Sommer (2015)] to the noisy case — while still a significant advancement in applied mathematics with actual practical implications — would not necessarily merit publication in PNAS on its own. It seems to us, however, that the reviewer hasn’t noticed that our main result in fact generalizes the noise-free result of [Hillar & Sommer (2015)] far beyond a simple extension to the noisy case in ways which provide greatly more valuable insight into the dictionary learning model, and that a significant deviation from their approach was necessary to prove these new facts. Moreover, our solution (Cor. 2) to the optimization formulation of the dictionary learning problem (Prob. 2) — that of most interest to optimization theorists and practitioners of their methods — is a significant result that is, to our knowledge, the first of its kind in both the noise-free and noisy domains. We discuss all of these points below in our response to the reviewer’s “detailed comments”.

Before moving on to our response to the detailed comments, we address briefly the issues of computability and tightness raised by the reviewer. The sufficient conditions for recovery are indeed computable and we have updated the manuscript to include a discussion of this fact. It is true, however, that we have not proven the tightness of these conditions and the manuscript is therefore not *conclusive* regarding the central problems of sparse linear coding. We believe that the fundamental characteristics of our results will likely remain unchanged upon any further tightening, and have opted instead to include in our revised manuscript a discussion of where such improvements would come from (e.g. a tightening of Lemma 4). Regarding the sample complexity N, we believe that we have uncovered the fundamental underlying combinatorial structure at play (i.e. the intersection semi-lattice of a hyper-graph and its transformation under the map \pi) and that our N is on the order of the tightest possible given a lack of any additional assumptions at all on the reproducing matrix B and codes \tilde x_i.

@@

-- The paper does a good job of motivating the general problem of sparse coding to a general audience. In particular, there is a good review of the areas in which dictionary learning arises.

-- I think the biggest potential issue for publication in PNAS is that the significance of the result needs more discussion. Specifically, the authors could address why going from the noise-free case to the noisy case is not incremental. PNAS specifically targets submissions that go beyond what could be published in a more specialized journal. Could the authors, for example, contrast the proof technique between the noisy and noise-free cases? Or, perhaps, the practical implications? 
-----------------

RESPONSE:

With all due respect to the reviewer, we believe the practical implications of a noisy result over a noiseless result are plainly evident: never has there been an experiment without noise and an unstable model is therefore useless in practice. It seems to us that the reviewer’s fundamental concern is more likely that our proof may amount to too straight-forward of an extension of the approach taken in [Hillar & Sommer (2015)] to satisfy the criterion for publication in PNAS. We believe this concern stems from two misconceptions of the reviewer: that our main result (Thm. 1) is merely an extension of that in [Hillar & Sommer (2015)] to the noisy case, and that our proof technique is only an incremental tweak on the approach taken in [Hillar & Sommer (2015)].

It is understandable from the presentation of our proof of our main result (Thm. 1) that our approach appears on the surface to be a straight-forward extension of that taken in [Hillar & Sommer (2015)]. As observed by our second reviewer, however, the extension to the noisy case did indeed require an original combination of several results in the literature. Aside from these required modifications, our proof of (Thm. 1) differs from the approach taken by [Hillar & Sommer (2015)] in a key way that is tucked into the proof of our Main Lemma — more specifically, by the addition of Lemma 5. [More detail on what goes down in Lemma 5?] We have updated the manuscript to make the reader more keenly aware of this. To be clear, the perspective we take on the problem yields the following powerful new conclusions beyond those from [Hillar & Sommer (2015)]:

@@

2) An extension to the case where the number of dictionary elements is unknown: The results of [Hillar & Sommer (2015)] only apply to the case where the matrix B has the same number of columns as A. We forgo this assumption and show that B must have at least as many columns as A and contains (up to noise) a subset of the columns of A. The size of this subset depends on a simple relation between the number of columns of B and the regularity of the support-set hypergraph.

3) No spark condition requirement for the generating matrix A: One of the mathematically significant contributions made by [Hillar & Sommer (2015)] was to forgo the constraint that the recovery matrix B also satisfy the spark condition, in contrast to all previously known uniqueness results which (either explicitly or implicitly) made this assumption. Our proof is powerful enough to show that, in fact, even the matrix A need not satisfy the spark condition to be identifiable! Rather, it need only be injective on the union of subspaces with supports that form sets in a hypergraph satisfying the SIP. We had omitted this point in our original submission of the manuscript to keep things simple, but have decided to include this interesting fact in our resubmission given that a recently approved NSF grant _____ makes use of this fact in its description of methods. This also required us to redefine the restricted matrix lower bound L to be in terms of a hypergraph, which is an interesting object for further study in its own right.

[4) Implications for the computational complexity of the dictionary learning problem: If the size of the support set of reconstructing codes is polynomial in m,k then the pigeonholing argument in the proof of Thm. 1 requires only a polynomial number of samples distributed over a polynomial number of supports; thus N is polynomial in m, k. [Chris? How does this have implications..?] This point was only hinted at in the Discussion section of our original submission, but we have included it in the updated manuscript to make clear the power of our approach over that taken in [Hillar & Sommer (2015)].

We also wish to reiterate that our solution to Problem 2, that of most interest to practitioners of dictionary learning methods, is the first of its kind in both the noise-free and noisy domains. We recognize that this was not clearly communicated in the Discussion section of our submitted manuscript, which begins: “In this note, we generalized the approach of (18) to prove the stability of unique solutions to Probs. 1 and 2 while significantly reducing the known sample complexity.” We have revised our re-submitted manuscript to make it clear that a solution to Problem 2 has never been supplied before in the literature until now [right?].

### 2017-08-03 manuscripts/PNAS/review/point_by_point_response.txt

Comments : 
SUMMARY OF THE PAPER

The paper finds conditions under which solutions for the sparse linear coding problem are unique. In sparse linear coding, one is given a matrix Y and the problem is to find a dictionary A and sparse coefficient matrix X such that Y=AX, up to a bounded uncertainty. This paper finds conditions under which A and X are unique up to scalings and permutation. Previous works looked at the noise-free case; this work considers the case of "noisy" measurements and proves a stability condition.

GENERAL COMMENTS

Overall, while I think that the manuscript makes a solid contribution, I believe is better suited to a more specialized journal. Unlike many journals, PNAS gives a criterion for significance: that the paper be within the top 10% of publications and should be of interest to a general scientific audience. Without compelling evidence to the contrary from the authors, it seems to me that the submission does not rise to that level. 
-----------------

@@

Comments : 
The paper presents an extension of existing conditions for uniqueness of the existence of a sparsity-inducing dictionary for a given set of data vectors that depends on the sampling of the resulting data subspaces. The extension considers the stability of dictionary learning in terms of the degree of variability present in distinct, feasible sparsity dictionary representations for the data under a constraint on the representation accuracy. A similar condition on the sampling of the underlying subspaces is obtained.

The paper's writing and composition is clear (with an exception detailed below), concise, and relevant to the topic. The analytical work appears to be correct to the best of my understanding, and the paper's combined use of several results in the literature to extend an existing dictionary learning uniqueness condition to a new dictionary learning stability condition appears to be original.

Some questions that remain and may be worth discussing deal with the practical impact of these results.

When is it possible to determine (numerically) if enough data has been gathered to obtain an accurate estimate of the underlying sparsity dictionary? 
-----------------
